### 2020-09-09 Manuscript/What.lies.ahead.reviewed.tex

Given the sea-surface temperatures and pH values for 2017 (observed) and 2095 (estimated) and taxon-specific logistic regression models, we then evaluated the suitability of habitat for each taxon in the future scenario. For each point in the pH - temperature grid, we calculated a species' probability of presence as the mean of 100 independent draws of the posterior model response. For richness estimations in future scenarios, we used each of the 100 draws of the posteriors in each environmental point as a sample and tallied the number of species detected on each draw. The mean value across the 100 draws was the input for a Wilcoxon test of differences in species richness between the 2017 and 2095 scenarios in each region. We performed the Wilcoxon test globally (total species richness) and on each phylum independently.

@@

These communities were distinguished by changes in the relative abundances of a wide variety of taxa. In the Hood Canal for example, the colder water and higher-pH community (community 1 in Fig. 2C) showed higher eDNA indexes of diatoms \textit{Thalassiosira nordenskioeldii}, \textit{Nitzschia cf. promare}, and \textit{Ditylum brightwellii} -- among other species -- relative to a higher-temperature community (community 3 in Fig. 2C). Communities 2 and 3 had relatively higher values for taxa from a mix of groups including \textit{Emiliana huxleyi}, \textit{Minutocellus polymorphus}, \textit{Micromonas pusilla}, and \textit{Bathycoccus prasinos}. Community 3 (Fig. 2A,B) and similar planktonic communities occupy the spectrum of environmental conditions most likely to be encountered in near-future climate scenarios as temperature rises and pH falls. For example, we expect the Hood Canal {\color{red} in 2095 to present the conditions in which community 3 is the most likely community 75\% of the time, an increase of 22\% compared to 2017(see Supplemental Material).} \st{T $>$ 12\degree C and pH $<$ {7.9} -- where community 3 is the most likely community -- more than four times as often in 2095 than in 2017 (4.9 \% vs 20.7\%).} {\color{blue} I couldn't find where you were doing this calculation, so I made one up -- see lines 1134 and following in the Rmd. BUT ALSO : community 3 isn't sensitive to pH, so what are we even saying here?}

@@

Given such heterogeneity in projections, gains and losses tend to balance one another out when looking at overall richness variation; {\color{red} we find no change in median richness for the year 2095 relative to the present in the Hood Canal (overall taxon richness by year, 95\% confidence interval in median species richness  -0.08, +0.04; Wilcoxon p $=$ 0.5); while higher diversity is expected in the San Juan Island in 2095 (increase in median species richness of 1.84-2.2, Wilcoxon p $< 10^{-16}$). Diatoms, however, show small but significant declines in richness in the Hood Canal (0.47-0.57 species, Wilcoxon p $< 10^{-16}$), while the changes on the San Juan Island are undetectable (median change 0-0.08 p $=$ 0.04).  Dinoflagellates see their richness increase in both regions with the future scenario(Wilcoxon p $< 10^{-16}$) }.  {\color{blue} I re-did some of these in the Rmd file (starting at line 1129, doing it line-by-line to convince myself what was going on. See line 1179 in particular: for diatoms, there is no change between years, but yet the wilcox p value is very low. I can't figure out what's happening there.}

%%%%%
 \begin{figure*}
 \centering
 \includegraphics[width=1\linewidth]{Figures_for_MS/Top_fig3_2020-09-03.Other.sel.png}
\caption{\footnotesize{Forecasted changes in plankton in the Salish Sea {\color{red} for Dinoflagellates (panel \textbf{A}), Diatoms (\textbf{B}) and Haptophytes (\textbf{C}). Each panels shows:  (\textbf{.1}) probability densities for the occurrence of selected taxa (species- and genus-level) for 2017 (blue) and 2095 (red); data are mean probabilities over 100 model draws, and variance in probability is due to differences in underlying environmental conditions.  (\textbf{.2}) Changes in relative species richness within each phylum across the simulated scenario. (\textbf{.3}) Relative taxon richness (raster color, warmer colors are more taxon-rich) for each of these same higher-taxa, for plausible ranges of pH and Temperature. Envelopes of observed (2017, blue) and modeled (2095, red) annual conditions in the Salish Sea shown for reference. Hood Canal and San Juan Island plotted separately to illustrate environmental differences between them.}}}

@@

{\color{blue} The paragraph below is important, and links to the question I had above: where are we doing this calculation in the code? I went through line by line; double check calculations -- see lines 1207 and following in the Rmd. Plus, a reviewer wanted a citation of the Coscinodiscus statement, so I added some. RYAN- I agree with the statement. Now we do these calculations in the script. I will add that the last bits of the prevalence of Alexandrium and Coscinoduscus are impactful and supercool}

{\color{red}The bulk of our projected community changes result from now-rare conditions occurring more frequently in the future. For example, in in the Hood Canal at present, we expect surface waters to have pH $<$ 7.9 and T $>$ 19$\degree$C only 1\% of the time. In 2095, we expect these conditions 6 times more frequently (i.e., 6\% of the time). At these values of T and pH, our model predicts the harmful \textit{Alexandrium} sp. to occur more often than not (mean frequency of occurrence $=$ 0.83). By contrast, the large centric diatom \textit{Coscinodiscus} -- a potentially key source of carbon for zooplankton and small fishes \cite{pasquaud2010determination, zamon2002tidal} with effects on dissolved oxygen and other water-column characteristics \cite{manabe1991bloom} -- occurs only one-third of the time under these same conditions (mean frequency $=$ 0.35).}

@@

 Our results therefore suggest a possible change in relative dominance between diatoms and other phytoplankton species such as dinoflagellates, consistent with those seen at ecological regime shifts found elsewhere \cite{wasmund2017diatom, hinder2012changes}. Such a shift could affect ecosystems in many ways; even under the assumption that the surviving taxa would maintain the primary production levels, for example, the smaller cell-size of dinoflagellates and the differential sinking rates of the two groups would likely alter regional patterns of nutrient cycling\cite{xiao2018warming, anderson2002harmful, bienfang1984sinking}. Although the north Atlantic has shown an increase in diatom abundance\cite{hinder2012changes}, the increase in wind stress and associated mixing in the water column in the open ocean is unlikely to occur in the Hood Canal, where stratification is the strongest in the Salish Sea \cite{moore2008descriptive}. Furthermore, locally focused models support an increase in dinoflagellate dominance with climate change, particularly during summer months \cite{khangaonkar2019salish}.

>>>>

Given the sea-surface temperatures and pH values for 2017 (observed) and 2095 (estimated) and taxon-specific logistic regression models, we then evaluated the suitability of habitat for each taxon in the future scenario. For each point in the pH - temperature grid, we calculated a species' probability of presence as the mean of 100 independent draws of the posterior model response. For each point, the sum of mean probabilities across species provided richness estimations. The mean value across the 100 draws was the input for a Wilcoxon test of differences in species richness between the 2017 and 2095 scenarios in each region. We performed the Wilcoxon test globally (total species richness) and on each phylum independently.

@@

These communities were distinguished by changes in the relative abundances of a wide variety of taxa. In the Hood Canal for example, the colder water and higher-pH community (community 1 in Fig. 2C) showed higher eDNA indexes of diatoms \textit{Thalassiosira nordenskioeldii}, \textit{Nitzschia cf. promare}, and \textit{Ditylum brightwellii} -- among other species -- relative to a higher-temperature community (community 3 in Fig. 2C). Communities 2 and 3 had relatively higher values for taxa from a mix of groups including \textit{Emiliana huxleyi}, \textit{Minutocellus polymorphus}, \textit{Micromonas pusilla}, and \textit{Bathycoccus prasinos}. Community 3 (Fig. 2A,B) and similar planktonic communities occupy the spectrum of environmental conditions most likely to be encountered in near-future climate scenarios as temperature rises and pH falls. For example, we expect the Hood Canal in 2095 to have the conditions in which community 3 is the most likely community 75\% of the time, an increase of 22\% compared to 2017(see Supplemental Material).

@@

Given such heterogeneity in projections, gains and losses tend to balance one another out when looking at overall richness variation; we find no change in median richness for the year 2095 relative to the present in the Hood Canal (overall taxon richness by year, 95\% confidence interval in median species richness  -0.08, +0.04; Wilcoxon p $=$ 0.5); while higher diversity is expected in the San Juan Island in 2095 (increase in median species richness of 1.84-2.2, p $< 10^{-16}$). Diatoms, in particular, show small but significant declines in richness in the Hood Canal (0.47-0.57 species, p $< 10^{-16}$), while the changes on the San Juan Island are negligible (median change 0-0.08 p $=$ 0.04).  Dinoflagellates see their richness increase in both regions with the future scenario (median change 0.14-0.18 Hood Canal; 0.21-0.25 San Juan Island; p $< 10^{-16}$ for each).

%%%%%
 \begin{figure*}
 \centering
 \includegraphics[width=1\linewidth]{Figures_for_MS/Top_fig3_2020-09-03.Other.sel.png}
\caption{\footnotesize{Forecasted changes in plankton in the Salish Sea for Dinoflagellates (panel \textbf{A}), Diatoms (\textbf{B}) and Haptophytes (\textbf{C}). Each panel shows:  (\textbf{1}) probability densities for the occurrence of selected taxa (species- and genus-level) for 2017 (blue) and 2095 (red); data are mean probabilities over 100 model draws, and variance in probability is due to differences in underlying environmental conditions.  (\textbf{2}) Changes in relative species richness within each phylum across the simulated scenario. (\textbf{3}) Relative taxon richness (raster color, warmer colors are more taxon-rich) for each of these same higher-taxa, for plausible ranges of pH and Temperature. Envelopes of observed (2017, blue) and modeled (2095, red) annual conditions in the Salish Sea shown for reference. Hood Canal and San Juan Island plotted separately to illustrate environmental differences between them.}}

@@

The bulk of our projected community changes result from now-rare conditions occurring more frequently in the future. For example, in in the Hood Canal at present, we expect surface waters to have pH $<$ 7.9 and T $>$ 19$\degree$C only 1\% of the time. In 2095, we expect these conditions 6 times more frequently (i.e., 6\% of the time). At these values of T and pH, our model predicts the harmful \textit{Alexandrium} sp. to occur more often than not (mean frequency of occurrence $=$ 0.83). By contrast, the large centric diatom \textit{Coscinodiscus} -- a potentially key source of carbon for zooplankton and small fishes \cite{pasquaud2010determination, zamon2002tidal} with effects on dissolved oxygen and other water-column characteristics \cite{manabe1991bloom} -- occurs only one-third of the time under these same conditions (mean frequency $=$ 0.35).

@@

 Our results therefore suggest a possible change in relative dominance between diatoms and other phytoplankton species such as dinoflagellates, consistent with those seen at ecological regime shifts found elsewhere \cite{wasmund2017diatom, hinder2012changes}. Such a shift could affect ecosystems in many ways; even under the assumption that the surviving taxa would maintain the primary production levels, for example, the smaller cell-size of dinoflagellates and the differential sinking rates of the two groups would likely alter regional patterns of nutrient cycling \cite{xiao2018warming, anderson2002harmful, bienfang1984sinking}. Although the north Atlantic has shown an increase in diatom abundance \cite{hinder2012changes}, the increase in wind stress and associated mixing in the water column in the open ocean is unlikely to occur in the Hood Canal, where stratification is the strongest in the Salish Sea \cite{moore2008descriptive}. Furthermore, locally focused models support an increase in dinoflagellate dominance with climate change, particularly during summer months \cite{khangaonkar2019salish}.
